---
title: "DATASUS Modules: SIM, SINASC, SIH, and SIA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DATASUS Modules: SIM, SINASC, SIH, and SIA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The `healthbR` package provides access to four DATASUS information systems via FTP, covering mortality, live births, hospital admissions, and outpatient production:

| Module | Function | Source document | Granularity | Years |
|--------|----------|-----------------|-------------|-------|
| SIM | `sim_data()` | Declaracao de Obito (DO) | Annual/UF | 1996--2024 |
| SINASC | `sinasc_data()` | Declaracao de Nascido Vivo (DN) | Annual/UF | 1996--2024 |
| SIH | `sih_data()` | AIH (Autorizacao de Internacao Hospitalar) | Monthly/UF | 2008--2024 |
| SIA | `sia_data()` | BPA / APAC | Monthly/type/UF | 2008--2024 |

All four modules share the same infrastructure:

- **DBC decompression**: .dbc files (compressed DBF) are decompressed internally using vendored C code -- no external dependencies required.
- **FTP download**: files are fetched from `ftp.datasus.gov.br` with automatic retry and exponential backoff.
- **Cache**: downloaded data is cached locally in Parquet (if `arrow` is installed) or .rds format.
- **Consistent API**: every module exposes `*_years()`, `*_info()`, `*_variables()`, `*_dictionary()`, `*_data()`, `*_cache_status()`, and `*_clear_cache()`.

## Getting started

```{r setup}
library(healthbR)
library(dplyr)
```

### Common helper functions

Each module provides the same set of helper functions. Here is a quick tour using SIM as an example:

```{r}
# available years
sim_years()
#> [1] 1996 1997 1998 ... 2023

# module information (data source, key variables, usage tips)
sim_info()

# list all variables with descriptions
sim_variables()

# search for a specific variable
sim_variables(search = "causa")

# data dictionary with category labels
sim_dictionary("SEXO")
```

The same pattern works for `sinasc_*()`, `sih_*()`, and `sia_*()`.

## SIM -- Mortality

The SIM (Sistema de Informacoes sobre Mortalidade) contains individual death records based on the Declaracao de Obito (DO).

### Basic download

```{r}
# all deaths in Acre, 2022
obitos_ac <- sim_data(year = 2022, uf = "AC")
obitos_ac
```

### Filter by cause of death

The `cause` parameter filters by underlying cause of death (CAUSABAS) using CID-10 prefix matching:

```{r}
# deaths from acute myocardial infarction (I21)
obitos_iam <- sim_data(year = 2022, uf = "AC", cause = "I21")

# all cardiovascular deaths (chapter I)
obitos_cardio <- sim_data(year = 2022, uf = "AC", cause = "I")
```

### Key variables

| Variable | Description |
|----------|-------------|
| `CAUSABAS` | Underlying cause of death (CID-10) |
| `DTOBITO` | Date of death |
| `SEXO` | Sex (M = Male, F = Female, I = Unknown) |
| `IDADE` | Age (encoded: 1st digit = unit, digits 2-3 = value) |
| `CODMUNRES` | Municipality of residence (IBGE code) |

### Example: deaths by cause chapter

```{r}
obitos_ac <- sim_data(year = 2022, uf = "AC")

obitos_ac |>
  mutate(chapter = substr(CAUSABAS, 1, 1)) |>
  count(chapter, sort = TRUE)
```

## SINASC -- Live births

The SINASC (Sistema de Informacoes sobre Nascidos Vivos) contains individual live birth records from the Declaracao de Nascido Vivo (DN).

### Basic download

```{r}
nasc_ac <- sinasc_data(year = 2022, uf = "AC")
nasc_ac
```

### Filter by congenital anomaly

The `anomaly` parameter filters by the CODANOMAL variable using CID-10 prefix matching:

```{r}
# births with any congenital anomaly (chapter Q)
anomalias <- sinasc_data(year = 2022, uf = "AC", anomaly = "Q")
```

### Key variables

| Variable | Description |
|----------|-------------|
| `DTNASC` | Date of birth |
| `SEXO` | Sex (1 = Male, 2 = Female, 0 = Unknown) |
| `PESO` | Birth weight (grams) |
| `IDADEMAE` | Mother's age |
| `CODMUNRES` | Municipality of residence (IBGE code) |
| `CODANOMAL` | Congenital anomaly code (CID-10) |

### Example: birth weight distribution

```{r}
nasc_ac <- sinasc_data(year = 2022, uf = "AC")

nasc_ac |>
  mutate(peso_num = as.numeric(PESO)) |>
  filter(!is.na(peso_num), peso_num > 0) |>
  mutate(weight_group = case_when(
    peso_num < 1500 ~ "Very low (<1500g)",
    peso_num < 2500 ~ "Low (1500-2499g)",
    peso_num < 4000 ~ "Normal (2500-3999g)",
    TRUE            ~ "High (>=4000g)"
  )) |>
  count(weight_group)
```

## SIH -- Hospital admissions

The SIH (Sistema de Informacoes Hospitalares) contains individual hospital admission records from the AIH (Autorizacao de Internacao Hospitalar). Unlike SIM and SINASC, data is organized **monthly**.

### Basic download

```{r}
# admissions in Acre, January 2022
intern_jan <- sih_data(year = 2022, month = 1, uf = "AC")
intern_jan
```

### The `month` parameter

SIH data is monthly -- one file per UF per month. Use `month` to control which months to download:

```{r}
# single month
sih_data(year = 2022, month = 6, uf = "AC")

# first semester
sih_data(year = 2022, month = 1:6, uf = "AC")

# all 12 months (default when month = NULL -- downloads 12 files per UF)
sih_data(year = 2022, uf = "AC")
```

### Filter by diagnosis

The `diagnosis` parameter filters by the principal diagnosis (DIAG_PRINC) using CID-10 prefix matching:

```{r}
# respiratory admissions (chapter J)
resp <- sih_data(year = 2022, month = 1, uf = "AC", diagnosis = "J")

# pneumonia specifically (J12-J18)
pneum <- sih_data(year = 2022, month = 1, uf = "AC",
                  diagnosis = c("J12", "J13", "J14", "J15", "J16", "J17", "J18"))
```

### Key variables

| Variable | Description |
|----------|-------------|
| `DIAG_PRINC` | Principal diagnosis (CID-10) |
| `DT_INTER` | Admission date |
| `SEXO` | Sex (1 = Male, 3 = Female, 0 = Unknown) |
| `MORTE` | In-hospital death (1 = Yes, 0 = No) |
| `VAL_TOT` | Total value (R$) |
| `DIAS_PERM` | Length of stay (days) |

### Example: admissions by diagnosis chapter

```{r}
intern <- sih_data(year = 2022, month = 1, uf = "AC")

intern |>
  mutate(chapter = substr(DIAG_PRINC, 1, 1)) |>
  count(chapter, sort = TRUE)
```

## SIA -- Outpatient production

The SIA (Sistema de Informacoes Ambulatoriais) contains outpatient production records. Like SIH, data is monthly, but SIA also has **13 file types** covering different categories of outpatient care.

### File types

| Code | Name | Description |
|------|------|-------------|
| PA | Producao Ambulatorial | BPA consolidated (default) |
| BI | Boletim Individualizado | BPA individualized |
| AD | APAC Laudos Diversos | High-complexity authorizations |
| AM | APAC Medicamentos | High-cost medications |
| AN | APAC Nefrologia | Nephrology procedures |
| AQ | APAC Quimioterapia | Oncology chemotherapy |
| AR | APAC Radioterapia | Oncology radiotherapy |
| AB | APAC Cirurgia Bariatrica | Bariatric surgery |
| ACF | APAC Confeccao de Fistula | Arteriovenous fistula |
| ATD | APAC Tratamento Dialitico | Dialysis |
| AMP | APAC Acompanhamento Multiprofissional | Multiprofessional follow-up |
| SAD | RAAS Atencao Domiciliar | Home care services |
| PS | RAAS Psicossocial | CAPS and psychosocial services |

### Basic download

```{r}
# outpatient production in Acre, January 2022 (default type = "PA")
ambul_jan <- sia_data(year = 2022, month = 1, uf = "AC")
ambul_jan

# different file type: high-cost medications
med <- sia_data(year = 2022, month = 1, uf = "AC", type = "AM")
```

### Filter by procedure and diagnosis

```{r}
# filter by SIGTAP procedure code (prefix match on PA_PROC_ID)
consult <- sia_data(year = 2022, month = 1, uf = "AC", procedure = "0301")

# filter by CID-10 diagnosis (prefix match on PA_CIDPRI)
resp <- sia_data(year = 2022, month = 1, uf = "AC", diagnosis = "J")
```

### Key variables (PA type)

| Variable | Description |
|----------|-------------|
| `PA_PROC_ID` | Procedure code (SIGTAP) |
| `PA_CIDPRI` | Principal diagnosis (CID-10) |
| `PA_SEXO` | Sex (1 = Male, 2 = Female) |
| `PA_IDADE` | Patient age |
| `PA_VALAPR` | Approved value (R$) |
| `PA_QTDAPR` | Approved quantity |

### Example: production by procedure group

```{r}
ambul <- sia_data(year = 2022, month = 1, uf = "AC")

ambul |>
  mutate(proc_group = substr(PA_PROC_ID, 1, 2)) |>
  count(proc_group, sort = TRUE)
```

## Cross-module analyses

A key strength of `healthbR` is the ability to combine data from different DATASUS modules and Census denominators in a single workflow. Below are three practical examples.

### Mortality rate (SIM + Census)

Calculate the crude cardiovascular mortality rate per 100,000 population:

```{r}
# step 1: count cardiovascular deaths in Sao Paulo, 2022
obitos_cardio <- sim_data(year = 2022, uf = "SP", cause = "I")
n_obitos <- nrow(obitos_cardio)

# step 2: get population denominator from Census 2022
pop_sp <- censo_populacao(year = 2022, territorial_level = "state") |>
  filter(grepl("Paulo", territorial_unit))

# step 3: calculate rate
taxa_mortalidade <- n_obitos / pop_sp$population * 100000
taxa_mortalidade
```

### Live births to deaths ratio (SINASC + SIM)

Compare the number of live births and deaths in a state:

```{r}
# births and deaths in Acre, 2022
nascimentos <- sinasc_data(year = 2022, uf = "AC")
obitos <- sim_data(year = 2022, uf = "AC")

razao <- nrow(nascimentos) / nrow(obitos)
razao
#> ratio > 1 means more births than deaths (population growth)
```

### Hospital vs. outpatient care (SIH + SIA)

Compare volumes and costs of respiratory care (CID-10 chapter J) between hospital and outpatient settings:

```{r}
# hospital admissions for respiratory diseases, January 2022
intern_resp <- sih_data(year = 2022, month = 1, uf = "AC", diagnosis = "J")

# outpatient production for respiratory diseases, January 2022
ambul_resp <- sia_data(year = 2022, month = 1, uf = "AC", diagnosis = "J")

# compare volumes
n_internacoes <- nrow(intern_resp)
n_ambulatorial <- nrow(ambul_resp)

# compare costs
custo_intern <- sum(as.numeric(intern_resp$VAL_TOT), na.rm = TRUE)
custo_ambul <- sum(as.numeric(ambul_resp$PA_VALAPR), na.rm = TRUE)

tibble::tibble(
  setting = c("Hospital (SIH)", "Outpatient (SIA)"),
  records = c(n_internacoes, n_ambulatorial),
  total_cost_brl = c(custo_intern, custo_ambul)
)
```

## Cache and performance

### Automatic caching

All DATASUS modules cache downloaded data automatically. When the `arrow` package is installed, data is saved in Parquet format (fast and compact); otherwise, .rds is used as fallback.

```{r}
# install arrow for optimized caching (recommended)
install.packages("arrow")
```

### Cache management

Each module provides `*_cache_status()` and `*_clear_cache()`:

```{r}
# check what is cached
sim_cache_status()
sih_cache_status()
sia_cache_status()

# clear cache for a specific module
sim_clear_cache()
```

### Tips for managing downloads

- **Use `uf`** to download only the states you need instead of all 27.
- **Use `month`** (SIH and SIA) to limit monthly downloads. Downloading a full year for all states requires 324 files per module (27 UFs x 12 months).
- **Use `vars`** to keep only the variables you need, reducing memory usage.
- SIM and SINASC are annual (one file per UF per year), so a full-year download is 27 files.
- SIH and SIA are monthly, so a full-year download is 324 files per type. SIA has 13 file types, so downloading everything for one year would be 4,212 files -- always filter by `type`, `uf`, and `month`.

## Additional resources

- [DATASUS TabNet](https://datasus.saude.gov.br/informacoes-de-saude-tabnet/) -- online tabulation tool for DATASUS data
- [DATASUS FTP](ftp://ftp.datasus.gov.br/dissemin/publicos/) -- public FTP server with raw data files
- [CID-10 (WHO ICD-10)](https://icd.who.int/browse10/2019/en) -- International Classification of Diseases, 10th revision
- [SIGTAP](http://sigtap.datasus.gov.br/) -- procedure code table for SUS (SIA/SIH)
