---
title: "Vaccination Data with SI-PNI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vaccination Data with SI-PNI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The **SI-PNI (Sistema de Informacao do Programa Nacional de Imunizacoes)** is Brazil's national immunization information system, managed by the Ministry of Health. It tracks vaccination doses applied and coverage rates across the country.

SI-PNI data come in two eras:

| Era | Years | Data type | Granularity |
|-----|-------|-----------|-------------|
| **Aggregated** | 1994--2019 | Dose counts (DPNI) and coverage (CPNI) | Annual per UF |
| **Microdata** | 2020+ | Individual-level (one row per dose) | Monthly |

`sipni_data()` automatically routes to the correct era based on the requested year.

## Data sources: the R2 mirror and DATASUS

By default, `healthbR` reads SI-PNI data from the
[healthbr-data](https://github.com/SidneyBissoli/healthbr-data) mirror:
hive-partitioned Parquet on Cloudflare R2 (free egress), with values
**byte-identical** to the Ministry's files and full provenance metadata
(source URL, hash, download date, pipeline version) embedded in every file.
If the mirror is unreachable, `sipni_data()` falls back automatically to the
official DATASUS/OpenDataSUS sources.

```{r}
# default: R2 mirror with automatic DATASUS fallback
sipni_data(year = 2024, uf = "AC", month = 1)

# pin a single source (no fallback)
sipni_data(year = 2019, uf = "AC", source = "datasus")

# invert the priority (DATASUS first, R2 as fallback)
sipni_data(year = 2019, uf = "AC", source = c("datasus", "r2"))
```

Why the mirror is the default:

- The Ministry decommissioned the old OpenDataSUS host in 2026 and removed
  the 2020--2025 microdata files from the new one; **only the mirror holds
  the complete 2020+ series**.
- Only the requested UF/month partitions are transferred (no ~1.4 GB
  national CSV download to extract one UF).
- Aggregates read in seconds (the DATASUS FTP is slow and unstable for
  large UFs).

The result records where the data actually came from:

```{r}
data <- sipni_data(year = 2024, uf = "AC", month = 1)
attr(data, "healthbr_source")
#>  microdata
#>       "r2"
attr(data, "healthbr_provenance")
#> # A tibble: 1 x 7  (partition, processing timestamp, Ministry source URL...)
```

### Checking availability with sipni_status()

The mirror publishes a `manifest.json` per dataset recording, for every
partition, the Ministry source file, its hash, the processing timestamp and
record counts. `sipni_status()` reads it:

```{r}
# everything the mirror holds
sipni_status()

# which 2026 microdata months are published so far?
sipni_status("microdados") |>
  filter(year == 2026)
```

### Check available years

```{r}
sipni_years()
#> [1] 1994 1995 ... 2025 2026
```

### Module information

```{r}
sipni_info()
```

## Aggregated data: doses applied (DPNI)

The default type returns aggregated dose counts (1994--2019):

```{r}
# doses applied in Acre, 2019
ac_doses <- sipni_data(year = 2019, uf = "AC")
ac_doses
```

### Key variables (DPNI)

| Variable | Description |
|----------|-------------|
| ANO | Reference year |
| UF | UF code (IBGE 2 digits) |
| MUNIC | Municipality code (IBGE 6 digits) |
| IMUNO | Immunobiological code |
| DOSE | Dose type (1st, 2nd, booster, etc.) |
| QT_DOSE | Number of doses applied |
| FX_ETARIA | Age group (coded) |

### Using the dictionary

By default the dictionary is read from the mirror — the full versions
converted from the Ministry's original .cnv/.dbf files.

**Important — decoding data requires `lookup = TRUE`.** In the .cnv files,
`code` is a sequential category code, and the value(s) actually found in
the data columns live in `source_codes` (possibly several per label,
reflecting code changes over the years — e.g. data codes `08` **and** `82`
both mean Hepatite B). Joining data against `code` silently produces wrong
labels. `lookup = TRUE` expands the dictionary into one row per **data
code**, ready to join:

```{r}
# published form (code = .cnv category, source_codes = data codes)
sipni_dictionary("IMUNO")

# join-ready lookup: one row per data code
sipni_dictionary("IMUNO", lookup = TRUE)

# dose types and age groups
sipni_dictionary("DOSE", lookup = TRUE)
sipni_dictionary("FX_ETARIA", lookup = TRUE)

# the built-in offline copy (already in data-code form)
sipni_dictionary("IMUNO", source = "datasus")
```

When more than one .cnv category claims the same data code, the more
specific claim wins: explicitly listed codes take precedence over codes
that only fall inside a range. This matters for residual catch-all
categories: FX_ETARIA's "Idade ignorada" spans codes `00-99`, but it only
labels codes that no specific age group claimed explicitly.

## Aggregated data: vaccination coverage (CPNI)

The CPNI type provides coverage rates per municipality:

```{r}
# vaccination coverage in Acre, 2019
ac_coverage <- sipni_data(year = 2019, type = "CPNI", uf = "AC")
ac_coverage
```

### Key variables (CPNI)

| Variable | Description |
|----------|-------------|
| ANO | Reference year |
| UF | UF code (IBGE 2 digits) |
| MUNIC | Municipality code (IBGE 6 digits) |
| IMUNO | Immunobiological code |
| QT_DOSE | Number of doses applied |
| POP | Target population |
| COBERT | Vaccination coverage (%) |

## Microdata (2020+)

For years 2020 and later, SI-PNI provides individual-level microdata
(one row per vaccination dose). The `type` parameter is ignored for these
years:

```{r}
# microdata for Acre, January 2024
ac_micro <- sipni_data(year = 2024, uf = "AC", month = 1)
ac_micro
```

**Column names differ by source.** The mirror publishes the Ministry's JSON
exports (56 fields, no CSV serialization artifacts); the OpenDataSUS CSVs
use different names (~47 fields). Each source returns its columns exactly
as published — healthbR does not rename or remap them:

| R2 mirror (default) | DATASUS CSV |
|---------------------|-------------|
| `dt_vacina` | `data_vacina` |
| `ds_vacina` | `descricao_vacina` |
| `tp_sexo_paciente` | `tipo_sexo_paciente` |
| `nu_idade_paciente` | `numero_idade_paciente` |
| `sg_uf_estabelecimento` | `sigla_uf_estabelecimento` |

### Key variables (R2 microdata)

| Variable | Description |
|----------|-------------|
| sg_uf_estabelecimento | UF of the health facility |
| co_municipio_estabelecimento | Municipality (IBGE) |
| tp_sexo_paciente | Sex (M/F) |
| nu_idade_paciente | Patient age |
| no_raca_cor_paciente | Race/color (descriptive) |
| ds_vacina | Vaccine name |
| ds_dose_vacina | Dose description |
| dt_vacina | Vaccination date |

### Exploring variables

```{r}
# DPNI variables
sipni_variables()

# CPNI variables
sipni_variables(type = "CPNI")

# microdata variables, R2 mirror (default; 56 fields)
sipni_variables(type = "API")

# microdata variables, OpenDataSUS CSV (~47 fields)
sipni_variables(type = "API", source = "datasus")

# search
sipni_variables(search = "dose")
```

## Month parameter for microdata

Each month is a separate partition. Use `month` to select specific months:

```{r}
# single month
jan <- sipni_data(year = 2024, uf = "AC", month = 1)

# first quarter
q1 <- sipni_data(year = 2024, uf = "AC", month = 1:3)

# all 12 months (default)
full_year <- sipni_data(year = 2024, uf = "AC")
```

For aggregated data (1994--2019), the `month` parameter is ignored because
the files are annual.

## Example: vaccine doses by immunobiological

```{r}
ac_2019 <- sipni_data(year = 2019, uf = "AC")

# decode immunobiological names: lookup = TRUE gives data-code rows
imuno_labels <- sipni_dictionary("IMUNO", lookup = TRUE) |>
  select(code, label)

doses_by_vaccine <- ac_2019 |>
  group_by(IMUNO) |>
  summarize(total_doses = sum(as.integer(QT_DOSE), na.rm = TRUE),
            .groups = "drop") |>
  left_join(imuno_labels, by = c("IMUNO" = "code")) |>
  arrange(desc(total_doses))

doses_by_vaccine
```

Note: the 2019 aggregated data are drastically incomplete *at the source*
(the Ministry was migrating to the new SI-PNI that year); totals are a
fraction of 2018's. This is faithful to what DATASUS publishes.

## Example: coverage trends over time

```{r}
# coverage data for Sao Paulo, 2015-2019
sp_cov <- sipni_data(
  year = 2015:2019,
  type = "CPNI",
  uf = "SP"
)

# average coverage by year
sp_cov |>
  group_by(year) |>
  summarize(
    mean_coverage = mean(as.numeric(COBERT), na.rm = TRUE),
    .groups = "drop"
  )
```

## Example: individual-level analysis (2020+)

```{r}
# vaccinations in Acre, January 2024
ac_jan <- sipni_data(year = 2024, uf = "AC", month = 1)

# vaccines administered
ac_jan |>
  count(ds_vacina, sort = TRUE)

# doses by sex
ac_jan |>
  count(tp_sexo_paciente)

# age distribution
ac_jan |>
  mutate(age = as.integer(nu_idade_paciente)) |>
  filter(!is.na(age)) |>
  mutate(age_group = cut(age,
                         breaks = c(0, 5, 12, 18, 30, 60, Inf),
                         right = FALSE)) |>
  count(age_group)
```

## Mixed year requests

When requesting years that span both eras (e.g., 2019 and 2024),
`sipni_data()` fetches each era and combines the results.
Note that column names and structure differ between eras:

```{r}
# aggregated (2019) + microdata (2024)
mixed <- sipni_data(year = c(2019, 2024), uf = "AC", month = 1)

# aggregated (UPPERCASE) and microdata columns are combined
# with NAs where columns don't overlap
names(mixed)
```

## Lazy evaluation over the mirror

With `lazy = TRUE` and the default source, `sipni_data()` returns the
**remote** arrow dataset — dplyr verbs are pushed down to R2 and only the
partitions your query touches are transferred. In this mode the partition
columns keep the bucket layout names (`ano`, `mes`, `uf`, as strings):

```{r}
ds <- sipni_data(year = 2019, uf = "AC", lazy = TRUE)
ds |>
  filter(IMUNO == "02") |>   # data code 02 = BCG (see sipni_dictionary)
  select(MUNIC, DOSE, QT_DOSE) |>
  collect()
```

## Download tips

- Aggregates (1994--2019) read from the mirror in seconds.
- Microdata months for large UFs (SP, MG) hold millions of rows — start
  with a single month and UF, or use `lazy = TRUE` and aggregate before
  `collect()`.
- Results are cached locally per partition: a second request for the same
  year/month/UF is served from disk.
- If you pin `source = "datasus"` for 2020+, only the current year's CSVs
  exist at the source, and each month is a ~1.4 GB national download.

## Smart type parsing

```{r}
# parsed types (default)
ac <- sipni_data(year = 2019, uf = "AC")
class(ac$QT_DOSE)  # integer

# raw character columns, exactly as published
ac_raw <- sipni_data(year = 2019, uf = "AC", parse = FALSE)
```

## Cache management

Downloaded data is cached locally for faster future access:

```{r}
# check cache status
sipni_cache_status()

# clear cache if needed
sipni_clear_cache()
```

## Additional resources

- [healthbr-data](https://github.com/SidneyBissoli/healthbr-data) — the
  mirror's pipelines, reproducibility policy and dataset documentation
- OpenDataSUS (`dadosabertos.saude.gov.br`)
- [Census vignette](censo-denominadores.html) for population denominators
